Waking up a sleeping rabbit:\\ On Natural-Language Sentence Generation with FF

### 2010-02-26 trunk/Papers/crisp-ff-10/conclusion.tex
intro umgeschrieben, platz gemacht.

Generation of natural language sentences is interesting as a planning
problem, both in its application value and in the challenges it
poses. After the changes described in this paper, FF is a practical
option for this application. It is striking how simple these changes
were, and certainly this suggests to look more closely at automatic
planner configuration techniques.

The domain is also far from ``solved'' in all its aspects. As we
pointed out, several features of the domain are very hard to capture
with existing search heuristics. In Fig.~\ref{fig:runtimes}, this
shows in the fact that no planner version scales beyond $n>4$ -- even
in case (a) where no adjectives at all are required. It remains an
interesting challenge to address this.

>>>>

Generation of natural language sentences is interesting as a planning
problem, both in its application value and in the challenges it
poses. After the changes to FF described in this paper, for the first
time an automated planner is a practical option for this
application. The simplicity of the changes required is striking, and
strongly suggests to look more closely at automatic planner
configuration techniques.

%It is striking how simple these changes were, and certainly this
%suggests to look more closely at automatic planner configuration
%techniques.

The domain is far from ``solved'' in all its aspects. As we pointed
out, several features of the domain are very hard to capture with
existing search heuristics. In Fig.~\ref{fig:runtimes}, this shows in
the fact that no planner version scales beyond $n>4$ -- even in case
(a) where no adjectives at all are required. It remains an interesting
challenge to address this.

### 2010-02-26 trunk/Papers/crisp-ff-10/crisp-ff.tex
intro umgeschrieben, platz gemacht.

FF exhibits a number of problems in this domain that were previously
unknown.  The basic observation of FF's inefficiency on these inputs
was made before
\cite{koller10:_exper_with_plann_for_natur_languag_gener}; here we
examine its causes in detail. For some of the observed pitfalls, we
are able to provide simple fixes. Others pose serious challenges to
planning research.

@@

FF's preprocessor contains two details that, before this paper, were
so detrimental that we were not able to seriously propose FF as a
practical solution.
%Simple fixes
%to these details are so effective that FF is now a practical (if not
%perfect) possibility.

@@

FF's preprocessor contains a very small imprudence that apparently
didn't surface in other domains, but that does in ours. FF
distinguishes operators into ``easy'' (DNF precondition) and ``hard''
(non-DNF precondition), which go through different instantiation
procedures. The procedure for hard operators uses a data structure
encompassing an integer for every possible instantiation of all
predicates. Since this does not take into account reachability, it is
prohibitively huge in our domain. Now, we actually have no hard
operators. But FF builds the data structure anyway, which exhausts
memory and causes a segmentation fault on serious instances.

@@

FF's preprocessing pitfalls -- and the fact that they didn't occur to
anyone before -- are certainly baffling, although easily fixed. We now
turn our attention to some of FF's search techniques, variations of
which are used in many planners and which are very much not easily
fixed.

@@

First, we consider Koehler and Hoffmann's
\shortcite{koehler:hoffmann:jair-00} ``reasonable goal orderings'',
variants and extensions of which are in prominent use today, e.g.\ in
LAMA \cite{richter:etal:aaai-08}. FF as used in the 2000 competition
approximates reasonable orders as a pre-process and then partitions
the goal set into a {\em goal agenda}, a sequence of goal
subsets. Each subset is posed to the planner in isolation. The final
plan is the concatenation of the plans for all entries.

@@

In our improved FF, the goal agenda is simply switched off. Having
disposed of goal orderings in that way, let us turn our attention to
the relaxed plan heuristic. %
%While the goal agenda meltdown may be more amusing than alarming, we
%next report that relaxed plan heuristics are not far from such a
%meltdown either. They are fatally 
This meets its Waterloo as well, due to the unusual lifecycle of facts
encoding substitution nodes and distractors. 
%In instances where these
%constructs are important for the construction of a valid sentence,
These facts are required to be false in the goal, and most of them are
initially false. However, any valid plan must make many of them true
at some intermediate time point (e.g.\ distractors arise depending on
the decisions taken by the planner). Hence, from the perspective of
the relaxed plan heuristic, from one state to the next whole classes
of goals suddenly appear ``out of the blue''.
%These goals are not
%accounted for at all in the preceding state.
%, and hence the relaxed plan is
%completely un-informative.

@@

More generally, from this example we immediately see that, in the
terms of Hoffmann \shortcite{hoffmann:jair-05}, the exit distance from
local minima under $h^+$ is unbounded. Also, we have unrecognized
dead-ends, i.e., dead-end states that have a relaxed plan. An example
is the initial state of a generation problem that it unsolvable
because not all distractors can be removed; its $h$ value will still
be $1$. So our domain is in the most difficult class of Hoffmann's
\shortcite{hoffmann:jair-05} ``planning domain taxonomy''. While it
shares this property with several other challenging domains, the
extremity of turning EHC into a breadth-first search is unheard of.

@@

Can we design heuristics that are better at dealing with the lifecycle
of substitution nodes and distractors? It is easy to see that the
problem of not noticing them in the initial state is present for most
known heuristics
\cite{helmert:icaps-04,richter:etal:aaai-08,karpas:domshlak-ijcai-09,helmert:domshlak:icaps-09,cai:etal:icaps-09}. The
two notable eceptions are the $h^m$ family, and abstraction
heuristics. We ran a few tests with a variant of the example problem
involving 10 rabbits.  The Graphplan estimate for the initial state is
$4$, i.e., $h^2$ is better than $h^+$ here (although still far from
the real goal distance of $14$). With merge\&shrink
\cite{helmert:etal:icaps07}, we obtained the estimate $11$ for the
initial state, which is much better. Still the search space with that
latter heuristic remained fairly large ($5200$ nodes vs.\ $13800$
nodes for blind search), and it is not clear how well the technique
will scale to larger instances and instances with a more complex
grammar. Certainly, interesting challenges remain.

@@

The only FF technique that appears to be clearly useful here is its
action pruning technique, the ``helpful actions''. While the relaxed
plan is often much too short, it appears to always contain at least
one of the actions that actually are useful. 
%
%As our experiments data
%will show, it certainly prunes the set of available actions quite
%effectively.
%
%For example, in the
%10-distractor test instance, $16875$ nodes are explored without
%helpful actions pruning; with pruning, this reduces to $3078$. For
%larger examples, the factor between the two numbers tends to become
%larger. 
%
%
%Since EHC is often completely inadequate, and helpful actions are
%quite effective, 
%
Given the above observations, we modified FF to directly run
best-first search with helpful actions (as
% regarding EHC and helpful actions, our
%modified FF directly runs best-first search with helpful actions (as
opposed to original FF, which switches to best-first with{\em out}
helpful actions if EHC fails).

>>>>

%FF exhibits a number of problems in this domain that were previously
%unknown.

The basic observation of FF's inefficiency in the NLG domain was made
before \cite{koller10:_exper_with_plann_for_natur_languag_gener}; here
we examine its causes in detail. For some of the observed pitfalls, we
are able to provide simple fixes. Others pose serious challenges to
planning research.

@@

%FF's preprocessor contains two details that, before this paper, were
%so detrimental that we were not able to seriously propose FF as a
%practical solution.
%Simple fixes
%to these details are so effective that FF is now a practical (if not
%perfect) possibility.

@@

FF's preprocessor contains a very small imprudence that apparently
didn't surface in other domains, but that does in ours. FF
distinguishes operators into ``easy'' (DNF precondition) and ``hard''
(non-DNF precondition), which go through different instantiation
procedures. The procedure for hard operators uses a data structure
encompassing an integer for every possible instantiation of all
predicates. Since this does not take into account reachability, it is
prohibitively huge in our domain. Now, {\em we actually have no hard
  operators}. But FF builds the data structure anyway, which exhausts
memory and causes a segmentation fault on serious instances.

@@

FF's preprocessing pitfalls are certainly baffling, but easily
fixed. We now turn our attention to some of FF's search techniques,
variations of which are used in many planners, and which are very much
not easily fixed. Our solution for the moment are simple configuration
changes. %; these give much better performance but do .

@@

First, we consider Koehler and Hoffmann's
\shortcite{koehler:hoffmann:jair-00} ``reasonable goal orderings'',
variants and extensions of which are in prominent use today, e.g.\ in
LAMA \cite{richter:etal:aaai-08}. FF as used in the 2000 competition
approximates reasonable orders as a preprocess and then partitions the
goal set into a {\em goal agenda}, a sequence of goal subsets. Each
subset is posed to the planner in isolation. The final plan is the
concatenation of the plans for all entries.

@@

In our improved FF, the goal agenda is simply switched off. Having
thus disposed of goal orderings, let us turn our attention to the
relaxed plan heuristic. %
%While the goal agenda meltdown may be more amusing than alarming, we
%next report that relaxed plan heuristics are not far from such a
%meltdown either. They are fatally 
This meets its Waterloo in the unusual life-cycle of facts encoding
substitution nodes and distractors.
%In instances where these
%constructs are important for the construction of a valid sentence,
These facts are required to be false in the goal, and most of them are
initially false. However, any valid plan must make many of them true
at some intermediate time point (e.g.\ distractors arise depending on
the decisions taken by the planner). Hence, from the perspective of
the relaxed plan heuristic, from one state to the next whole classes
of goals suddenly appear ``out of the blue''.
%These goals are not
%accounted for at all in the preceding state.
%, and hence the relaxed plan is
%completely un-informative.

@@

More generally, from this example we immediately see that, in the
terms of Hoffmann \shortcite{hoffmann:jair-05}, the exit distance from
local minima under $h^+$ is unbounded. Also, we have unrecognized
dead-ends, i.e., dead-end states that have a relaxed plan. An example
is the initial state of a task that it unsolvable because not all
distractors can be removed; its $h$ value will still be $1$. So the
domain is in the most difficult class of Hoffmann's
\shortcite{hoffmann:jair-05} planning domain taxonomy. While it shares
this property with several other challenging domains, turning EHC into
a breadth-first search is unheard of.

@@

Can we design heuristics that are better at dealing with the life-cycle
of substitution nodes and distractors? It is easy to see that the
problem of not noticing them in the initial state is present for most
known heuristics
\cite{helmert:icaps-04,richter:etal:aaai-08,karpas:domshlak-ijcai-09,helmert:domshlak:icaps-09,cai:etal:icaps-09}. The
two notable exceptions are the $h^m$ family, and abstraction
heuristics. We ran a few tests with a variant of the example problem
involving 10 rabbits.  The Graphplan estimate for the initial state is
$4$, i.e., $h^2$ is better than $h^+$ here (although still far from
the real goal distance of $14$). With merge\&shrink
\cite{helmert:etal:icaps07}, we obtained the estimate $11$ for the
initial state, which is much better. Still the search space with that
latter heuristic remained fairly large ($5200$ nodes vs.\ $13800$
nodes for blind search), and it is not clear how well the technique
will scale to larger instances and instances with a more complex
grammar. Certainly, interesting challenges remain.

@@

The only FF technique that appears to be clearly useful here is its
action pruning technique. While the relaxed plan is often much too
short, it appears to always contain at least one of the actions that
actually are useful.
%
%As our experiments data
%will show, it certainly prunes the set of available actions quite
%effectively.
%
%For example, in the
%10-distractor test instance, $16875$ nodes are explored without
%helpful actions pruning; with pruning, this reduces to $3078$. For
%larger examples, the factor between the two numbers tends to become
%larger. 
%
%
%Since EHC is often completely inadequate, and helpful actions are
%quite effective, 
%
Given the above observations, we modified FF to directly run
best-first search with helpful actions (as
% regarding EHC and helpful actions, our
%modified FF directly runs best-first search with helpful actions (as
opposed to original FF, which switches to best-first with{\em out}
helpful actions if EHC fails).

### 2010-02-26 trunk/Papers/crisp-ff-10/introduction.tex
intro umgeschrieben, platz gemacht.

Existing planners are relatively good at solving modest instances of
the planning problems that arise in this context. However, Koller and
Petrick \shortcite{koller10:_exper_with_plann_for_natur_languag_gener}
recently showed that as the planning problem instances from Koller and
Stone \shortcite{KolSto07} scale up, off-the-shelf planners become too
slow to be practically useful. They noted that there are major sources
of inefficiency both in the search itself and in the commonly used
preprocesses, due to several features of the domain
(e.g.\ comparatively large predicate and operator arities) that
diverge from most IPC benchmarks.
%the sheer number of
%ground instances of facts and actions that come up in this domain.

In this paper, we revisit this situation, analyze the sources of
inefficiency in more detail in the context of the FF planner
\cite{HoffmannNebel01}, and show how to modify FF to obtain
dramatically better performance. We achieve this by tweaking the
implementation of FF's preprocessor, and by acting on some simple but
shocking observations regarding the behavior of some of FF's search
techniques, in particular its goal ordering technique and the relaxed
plan heuristic. The changes we make are all simple, but make FF
perform well enough to be useful in practical NLG applications.

@@

Our performance gains with simple FF configuration changes certainly
point out that automatic planner configuration may be quite
useful. Apart from that, our domain remains an interesting challenge
for the development of better heuristics. There still are realistic
instances of the NLG planning domain on which our modified FF is not
practical. Overcoming this weakness would entail overcoming the
observed weaknesses of FF's search techniques -- variants of which are
used in many planners today.

>>>>

Existing planners are relatively good at solving modest instances of
the planning problems that arise in this context. However, Koller and
Petrick \shortcite{koller10:_exper_with_plann_for_natur_languag_gener}
recently showed that as the planning problem instances from Koller and
Stone \shortcite{KolSto07} scale up, off-the-shelf planners become
slow -- indeed, {\em too slow to be practically useful in NLG
  applications.} They showed this for SGplan and several variants of
FF \cite{HoffmannNebel01}; we show it herein also for LAMA
\cite{richter:etal:aaai-08}. Koller and Petrick noted that there are
major sources of inefficiency both in the search itself and in the
commonly used preprocesses; they identified several features of the
domain (e.g.\ comparatively large predicate and operator arities) that
diverge from most IPC benchmarks.
%the sheer number of
%ground instances of facts and actions that come up in this domain.

In this paper, we revisit this situation, and analyze the sources of
inefficiency in detail for FF. Based on the analysis, we modify FF and
obtain {\em dramatically better performance that, for the first time,
  makes automated planners a realistic possibility for NLG
  applications.} To achieve this quantum leap, all that is needed are
a few minor implementation fixes and basic reconfiguration of search
options. While these fixes certainly are no contribution to planning
in their own right, we believe it a valuable -- and slightly
unsettling -- lesson learned that such banalities can make the
difference between success and failure. They can hold up progress in
planning applications for years (until a planning expert comes by and
does some code profiling). We interpret this as a strong call for
automated configuration techniques, e.g.\ along the lines of Hutter et
al.\ \shortcite{HutHooStu07}.

Apart from these insights, our domain remains an interesting challenge
for the development of better heuristics. There still are realistic
instances of the NLG planning domain on which our modified FF is not
practical. Overcoming this weakness would entail overcoming
fundamental weaknesses of FF's goal ordering techniques and the
relaxed plan heuristic -- variants of which are used in many planners
today.

% show how to modify FF to obtain dramatically better performance. We
% achieve this by tweaking the implementation of FF's preprocessor, and
% by acting on some simple but shocking observations regarding the
% behavior of some of FF's search techniques, in particular its goal
% ordering technique and the relaxed plan heuristic. The changes we
% make are all simple, but make FF perform well enough to be useful in
% practical NLG applications.

@@

%Our performance gains with simple FF configuration changes certainly
%point out that automatic planner configuration may be quite
%useful. Apart from that, our domain remains an interesting challenge
%for the development of better heuristics. There still are realistic
%instances of the NLG planning domain on which our modified FF is not
%practical. Overcoming this weakness would entail overcoming the
%observed weaknesses of FF's search techniques -- variants of which are
%used in many planners today.
